R for Clinical Study Reports and Submission

### 2021-11-01 tlf-overview.Rmd
fix grammar, trailing spaces, file organization

# Overview

@@

In clinical trials, a critical step is
submission to regulatory agencies. 
[Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
has become a worldwide regulatory submission standard format. 
For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd). Clinical Data Interchange Standards Consortium (CDISC) provided a 
[pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).

Within eCTD, clinical study reports (CSRs) are located at module 5. 
[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides 
compilation of structure and content of clinical study reports.

A typical CSR contains full details on the methods and results of an individual clinical study.
In support of the statistical anlaysis, a large number of tables, listings and figures are
incorporated into the main text and appendices.
In CDISC pilot project, an 
[example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf) 
is also provided. If you are interested in more examples of clinical study reports, 
the European Medicines Agency (EMA) publishes clinical data including 
clinical study reports submitted by pharmaceutical companies 
on [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).

Building CSRs is a teamwork between clinicians, medical writers, statisticians, statistical programmers
and other relevant specialists such as experts on biomarkers. 
Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
In an organization, they commonly work together to 
define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR to
summarize the safety and/or safety of the pharmaceutical product.
Microsoft Word is widely used to prepare CSR in the pharmaceutical industry. 
Therefore, `rtf`, `doc`, `docx` are commonly used formats in their deliverables.

In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in RTF format
that is commonly used in a CSR. The examples are in compliance of 
[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)

> FDA's PDF specification is a general reference, each organization can define 
> more specific TLF format requirement that can be different from the examples in this book.

@@

In the rest of this chapter, we are following the 
[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
on structure and content of clinical study reports.

@@

- Section 10: Study participants 
- Section 11: Efficacy evaluation
- Section 12: Safety evaluation 
- Section 14: Tables, listings and figures referrals but not included in the text
- Section 16: Appendices

@@

We used publicly available CDISC pilot 
[study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

For simplicity, we have downloaded all these datasets into `adam_data` folder of 
this project and transfer them from `.xpt` format to `sas7bdat` format.

@@

- [tidyverse](https://www.tidyverse.org/): prepare reporting ready datasets. 
- [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs

@@

RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/), 
and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.

There are also books to introduce tidyverse. 
We assume reader to have experience in using `tidyverse` in this book.

@@

- provide simple "verb" functions that correspond to each component of a table, 
  to help you translate a data frame to a table in RTF file.
- enable pipes (`%>%`). 
- only focus on the **table format**. 
  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)

Before creating an RTF table we need to:

- Figure out table layout. 
- Split the layout into small tasks in the form of a computer program.
- Execute the program.

We provided a brief introduction of `r2rtf` and show how to transfer 
data frames into Table, Listing, and Figures (TLFs).

Other extended examples and features are covered in the 
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

To explore the basic RTF generation verbs in `r2rtf`, 
we'll use the dataset `r2rtf_adae` saved in the `r2rtf` package. 
This dataset contains adverse events (AE) information from a clinical trial.

Below is the meaning of relevant variables. 
More information can be found in the help page of the dataset (`?r2rtf_adae`)

In this example we consider three variables:

@@

- `rtf_page`: RTF page information
- `rtf_title`: RTF title information
- `rtf_colheader`: RTF column header information
- `rtf_body`: RTF table body information
- `rtf_footnote`: RTF footnote information
- `rtf_source`: RTF data source information

All these verbs are designed to enable the usage of pipes (`%>%`).
A full list of all functions can be found in the 
[r2rtf package reference](https://merck.github.io/r2rtf/reference/index.html)

@@

- `rtf_body` is used to define table body layout. 
- `rtf_encode` transfers table layout information into RTF syntax.
- `write_rtf` save RTF encoding into a file with file extension `.rtf`

@@

If we want to adjust the width of each column to 
provide more space to the first column, 
this can be achieved by updating the `col_rel_width` argument 
in the `rtf_body` function.

@@

In this example, the defined relative width is `3:2:2:2`. 
Only the ratio of `col_rel_width` is used. 
Therefore it is equivalent to use `col_rel_width = c(6,4,4,4)` 
or `col_rel_width = c(1.5,1,1,1)`.

@@

In the previous example, we found the issue of a misaligned column header. 
We can fix the issue by using the `rtf_colheader` function.

In `rtf_colheader`, `colheader` argument is used to provide the content of the column header. 
We use `"|"` to separate the columns.

@@

In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, 
`text_justification` argument is used to align text. 
Default is `"c"` for center justification. 
To vary text justification by column, use character vector with length of vector equal to
number of columns displayed (e.g., `c("c","l","r")`).

@@

In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, etc.,
`border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.

If we want to remove the top border of `"Adverse Events"` in the header, 
we can change the default value `"single"` to `""` in the `border_top` argument as showed below.

@@

In this example, we also demonstrate the possibility to add multiple column headers.

@@

In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
there are more examples to illustrate how to customize

>>>>

# Overview {#tlf-overview}

@@

In clinical trials, a critical step is submission to regulatory agencies.
[Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
has become a worldwide regulatory submission standard format.
For example, the United States Food and Drug Administration (US FDA) required
new drug applications and biologics license applications
[must be submitted using the eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
The Clinical Data Interchange Standards Consortium (CDISC) provided a
[pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).

Within eCTD, clinical study reports (CSRs) are located at module 5.
[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides
a compilation of the structure and content of clinical study reports.

A typical CSR contains full details on the methods and results of an individual clinical study.
In support of the statistical analysis, a large number of tables, listings, and figures are
incorporated into the main text and appendices.
In the CDISC pilot project, an
[example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
is also provided. If you are interested in more examples of clinical study reports,
the European Medicines Agency (EMA) publishes clinical data, including
clinical study reports submitted by pharmaceutical companies
on the [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).

Building CSRs is teamwork between clinicians, medical writers, statisticians, statistical programmers,
and other relevant specialists such as experts on biomarkers.
Here, we focus on the work and deliverables completed by statisticians and statistical programmers.
In an organization, they commonly work together to
define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR to
summarize the safety and/or safety of the pharmaceutical product.
Microsoft Word is widely used to prepare CSR in the pharmaceutical industry.
Therefore, `rtf`, `doc`, `docx` are commonly used formats in their deliverables.

In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in RTF format
that is commonly used in a CSR. The examples are in compliance with the
[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download).

> FDA's PDF specification is a general reference. Each organization can define
> more specific TLF format requirements that can be different from the examples in this book.

@@

In the rest of this chapter, we are following the
[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
on the structure and content of clinical study reports.

@@

- Section 10: Study participants
- Section 11: Efficacy evaluation
- Section 12: Safety evaluation
- Section 14: Tables, listings, and figures referrals but not included in the text
- Section 16: Appendices

@@

We used publicly available CDISC pilot
[study data located in the CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

For simplicity, we have downloaded all these datasets into the `adam_data` folder of
this project and converted them from the `.xpt` format to the `.sas7bdat` format.

@@

- [tidyverse](https://www.tidyverse.org/): prepare datasets ready for reporting.
- [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs

@@

RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.

There are also books to introduce tidyverse.
We assume the reader to have experience in using `tidyverse` in this book.

@@

- provide simple "verb" functions that correspond to each component of a table,
  to help you translate a data frame to a table in an RTF file;
- enable pipes (`%>%`);
- focus on the **table format** only.
  Data manipulation and analysis shall be handled by other R packages (e.g., `tidyverse`).

Before creating an RTF table, we need to

- figure out the table layout;
- split the layout into small tasks in the form of a computer program;
- execute the program.

We provided a brief introduction of `r2rtf` and showed how to transfer
data frames into Table, Listing, and Figures (TLFs).

Other extended examples and features are covered on the
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

To explore the basic RTF generation verbs in `r2rtf`,
we will use the dataset `r2rtf_adae` saved in the `r2rtf` package.
This dataset contains adverse events (AE) information from a clinical trial.

We will begin by loading the packages:

```{r}
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
```

Below is the meaning of relevant variables.
More information can be found on the help page of the dataset (`?r2rtf_adae`)

In this example, we consider three variables:

@@

- `rtf_page()`: RTF page information
- `rtf_title()`: RTF title information
- `rtf_colheader()`: RTF column header information
- `rtf_body()`: RTF table body information
- `rtf_footnote()`: RTF footnote information
- `rtf_source()`: RTF data source information

All these verbs are designed to enable the usage of pipes (`%>%`).
A full list of all functions can be found in the
[r2rtf package function reference manual](https://merck.github.io/r2rtf/reference/index.html).

@@

- `rtf_body()` is used to define table body layout.
- `rtf_encode()` transfers table layout information into RTF syntax.
- `write_rtf()` save RTF encoding into a file with file extension `.rtf`

@@

If we want to adjust the width of each column to
provide more space to the first column,
this can be achieved by updating the `col_rel_width` argument
in the `rtf_body()` function.

@@

In this example, the defined relative width is `3:2:2:2`.
Only the ratio of `col_rel_width` is used.
Therefore it is equivalent to use `col_rel_width = c(6, 4, 4, 4)`
or `col_rel_width = c(1.5, 1, 1, 1)`.

@@

In the previous example, we found the issue of a misaligned column header.
We can fix the issue by using the `rtf_colheader()` function.

In `rtf_colheader()`, the `colheader` argument is used to provide the content of the column header.
We use `"|"` to separate the columns.

@@

In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`,
the `text_justification` argument is used to align text.
Default is `"c"` for center justification.
To vary text justification by column, use character vector with length of vector equal to
number of columns displayed (e.g., `c("c", "l", "r")`).

@@

In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`, etc.,
`border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.

If we want to remove the top border of `"Adverse Events"` in the header,
we can change the default value `"single"` to `""` in the `border_top` argument, as shown below.

@@

In this example, we also demonstrate the possibility of adding multiple column headers.

@@

In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
there are more examples to illustrate how to customize

### 2021-11-01 tlf-population.Rmd
fix grammar, trailing spaces, file organization

In this chapter, we illustrate how to create a summary table for
analysis population in a study.

@@

Before we write analysis code, let's discuss the possibility to reuse R
code by writing some toy helper functions.

@@

In Chapter \@ref(disposition), there are few repeating steps to:

@@

To calculate the numbers and percentages of participants by groups, we
provide a simple (but not robust) wrapper function, `count_by()` using
the `dplyr` and `tidyr` package.

@@

- Step 1: Read data into R, i.e., `adsl`.
- Step 2: Bind the counts/percentages of the ITT population, the
  efficacy population and the safety population by row using the
  `count_by()` function.
- Step 3: Format the output in Step 2 by `r2rtf`.

>>>>

In this chapter, we illustrate how to create a summary table for
the analysis population in a study.

@@

Before we write the analysis code, let's discuss the possibility of reusing R
code by writing some toy helper functions.

@@

In Chapter \@ref(disposition), there are a few repeating steps to:

@@

To calculate the numbers and percentages of participants by groups, we
provide a simple (but not robust) wrapper function, `count_by()`, using
the `dplyr` and `tidyr` package.

@@

- Step 1: Read data into R, i.e., `adsl`.
- Step 2: Bind the counts/percentages of the ITT population, the
  efficacy population, and the safety population by row using the
  `count_by()` function.
- Step 3: Format the output in Step 2 by `r2rtf`.
